Supplement to ``{\mytitle}''

### 2022-12-19 si/si.Rnw
Refactored files, addressed some of Ed's comments

<<packages, include=FALSE, cache=FALSE, echo=FALSE>>=
library(tidyverse)
library(pomp)
library(haitipkg)
library(spatPomp)
library(doParallel)
library(doRNG)

cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
# if(is.na(cores)) cores <- detectCores()
cores <- 20
registerDoParallel(cores)
theme_set(theme_bw())
@

<<set-opts, include=FALSE, cache=FALSE, echo=FALSE>>=

options(
  scipen = 2,
  help_type = "html",
  stringsAsFactors = FALSE,
  # prompt = "R> ",
  continue = "+  ",
  width = 70,
  useFancyQuotes = FALSE,
  reindent.spaces = 2,
  xtable.comment = FALSE
)
@

<<knitr-opts, include=FALSE, cache=FALSE, purl=FALSE>>=
library("knitr")
opts_knit$set(concordance=TRUE)
opts_chunk$set(
  progress=TRUE,prompt=TRUE,highlight=FALSE,
  tidy=TRUE,
  tidy.opts=list(
    keep.blank.line=FALSE
  ),
  comment="",
  warning=FALSE,
  message=FALSE,
  error=TRUE,
  echo=TRUE,
  cache=FALSE,
  strip.white=TRUE,
  results="markup",
  background="#FFFFFF00",
  size="normalsize",
  fig.path="figure/",
  fig.lp="fig:",
  fig.align="center",
  fig.show="asis",
  #    dpi=300,
  dev="pdf",
  dev.args=list(
    bg="transparent",
    pointsize=9
  )
)

@@

\newcommand\mytitle{Informing policy via dynamic models: Eliminating cholera in Haiti}
\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}

@@

In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
For a discrete population model, these rates define a Markov chain.
For a continuous and deterministic model, the rates define a system of ordinary differential equations.
Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
Our treatment follows \citet{breto09}.

A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments.
The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$, together with flows into and out of each compartment denoted by $N_{\bullet i}(t)$ and $N_{i\bullet}(t)$ respectively.
These flows are required to satisfy a ``conservation of mass'' identity:
\begin{equation}
X_i(t)=X_i(0)+N_{\bullet i}(t) - N_{i\bullet}(t) + \sum_{j\neq
i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
\end{equation}
Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\bullet$.

@@

\subsection{Simple Markov counting system interpretation}
\label{subsec:smcs}
A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
A basic approach to this is to define
\begin{eqnarray}
\label{eq:smcs1}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{ij}\big(t,X(t)\big) + o(\delta),
\\
\label{eq:smcs2}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
 &=& \delta \mu_{ij}\big(t,X(t)\big) + o(\delta),
\end{eqnarray}
for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\bullet\}$ with $i\neq j$.
As with the ODE case, we need special attention for flows into the system, and we define
\begin{eqnarray}
\label{eq:smcs3}
\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta),
\\
\label{eq:smcs4}
\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=1 \given X(t)\big]
 &=& \delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta).
\end{eqnarray}
Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
We therefore call the Markov chain a simple Markov counting system (SMCS).
The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.

@@

Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
Specifically, we consider a model given by
\begin{equation}
\label{eq:odmcs1}
\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
\end{equation}
where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
\begin{equation}
\label{eq:odmcs2}
\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
\end{equation}
Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by defining the solution of \myeqref{eq:odmcs1} to be the limit of an Euler scheme.
Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneus jumps of size greater than one \citep{breto11}.

@@

For the case $i=\bullet$, one can use
\begin{equation}
\label{eq:numerics6}
\Delta N_{\bullet j} \sim \mathrm{poisson}( \mu_{\bullet j} \delta),
\end{equation}
an independent Poisson random variable with mean $\mu_{\bullet j} \delta$, as was done in Model~1.

Another common approach is to balance the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\bullet i}(t) = \sum_{i}N_{i \bullet}(t)$, in order to make the model consistent with the known total population, as was done in Models~2 and~3. In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\bullet$ into $S$ to describe only net population increase.

@@

This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
This shift was justified by a "change of the case definition that occurred on January 1st, 2018";
this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
We therefore do not allow a change in reporting rate when fitting Model~3.

@@

To perform inference on POMP models, it is necessary to propose an initial density for the latent process $f_{X_0}(x_0;\theta)$.
This density is used to obtain initial values of the latent state when fitting and evaluating the model.
For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
It is also sometimes necessary to fit some initial value parameters in order to help determine starting values for weakly identifiable compartments.
In the following subsections, we mention initial value parameters that were fit for each model.

@@

For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I(0)$ and $E(0)$, respectively) using the MIF2 algorithm.
Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S(0) = 1 - I(0) - E(0)$.

@@

Model~2 assumes that the initial vales are known and constant, and so no initial value parameters need to be estimated.
Starting values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
Specifically, the model sets $I_{u}(0) = y^*_{u}(0) / \reportRate$ for each unit $u \in 1:10$, where $I_{u}(t)$ is the number of infected individuals in unit $u$ at time $t$, $y^*_{u}(t)$ is the reported number of cases, and $\reportRate$ is the reporting rate.

@@

As described in the main test, the latent states are initialized by estimating of the number of infected, asymptomatic, and recovered individuals at time $t_0$ by using observed case counts at time $t_1$.
The initial value for the bacteria compartment is then determined using the rainfall dynamics of Eq.~23.
For the departments of Grande'Anse, Nord-Ouest, and Sud-Est, estimating the latent states at time $t_0$ using the observed data at time $t_1$ is problematic (see Sec.~\ref{sec:mod3Cal}), so we instead estimate initial value parameters $I_{u0}(0)$ for the integers $u \in \{3, 7, 9\}$, which correspond to the aforementioned departments, and the remaining hidden states are estimated by enforcing model dynamics using this result.

@@

<<CalibrateOrig3, echo=FALSE>>=
set.seed(8636712)

# Create vectors for the unit and shared parameters
unit_specific_names <- c("betaB", "foi_add")
shared_param_names <- c(
  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
  "epsilon", "k"
)
est_param_names <- c(
  unit_specific_names, shared_param_names
)

@@

set_rw_1 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.01)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.01)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.01)
  } else {
    return(0)
  }
}

set_rw_2 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.003)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.003)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.003)
  } else {
    return(0)
  }
}

set_rw_3 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.0025)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.0025)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.0025)
  } else {
    return(0)
  }
}

@@

# Search hyper-parameters for each run-level and search
SEARCH1 = list(
  NBPF = 100,
  NP = 1000,
  SPAT_REGRESSION = 0.05,
  NREPS = 72,
  NP_EVAL = 1000,
  NREPS_EVAL = 8,
  RW_SD = chol_rw_1.sd,
  COOLING = 0.5,
  KEEP_TRACES = TRUE,
  KEEP_LIK_MAT = TRUE
)

SEARCH2 = list(
  TOP_N = 9,
  NBPF = 100,
  NP = 1000,
  SPAT_REGRESSION = 0.05,
  NREPS = 4,
  NP_EVAL = 1000,
  NREPS_EVAL = 9,
  RW_SD = chol_rw_2.sd,
  COOLING = 0.5,
  KEEP_TRACES = TRUE,
  KEEP_LIK_MAT = TRUE
)

SEARCH3 = list(
  TOP_N = 6,
  NBPF = 100,
  NP = 1000,
  SPAT_REGRESSION = 0.05,
  NREPS = 6,
  NP_EVAL = 1000,
  NREPS_EVAL = 9,
  RW_SD = chol_rw_3.sd,
  COOLING = 0.5,
  KEEP_TRACES = TRUE,
  KEEP_LIK_MAT = TRUE
)

@@

# TODO: Move to run_level_3 folder...

save_file <- "../model3/run_level_2/fit_orig_model.rds"

haiti3_orig_fit <- bake(
  file = save_file, {
    fit_coupled_haiti3(
      nsearches = N_SEARCHES,
      search1 = SEARCH1,
      search2 = SEARCH2,
      search3 = SEARCH3,
      search_rho = FALSE,
      search_gamma = FALSE,
      ncores = cores
    )
  }
)

best_m <- which.max(haiti3_orig_fit$search3$logLiks$logLik)

best_params <- haiti3_orig_fit$search3$params[best_m, ]

all_liks <- as.data.frame(haiti3_orig_fit$search3$likMat)
all_liks$which <- rep(1:36, each = 9)
best_mod <- filter(all_liks, which == best_m)

unitLikesM3 <- best_mod %>%
  pivot_longer(
    cols = -which,
    names_to = "unit",
    names_prefix = "V",
    values_to = "logLik"
  ) %>%
  select(-which) %>%
  group_by(unit) %>%
  summarize(
    ll_est = logmeanexp(logLik),
    ll_se = logmeanexp(logLik, se = TRUE)[2]
  ) %>%
  ungroup() %>%
  mutate(unit = as.numeric(unit)) %>%
  arrange(unit)
@

<<M3_bench, echo=FALSE, message=FALSE, include=FALSE>>=
MODEL3_CASES <- haitiCholera %>%
  dplyr::rename(
    date = date_saturday, Grande_Anse = Grand.Anse,
    Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
    Sud_Est = Sud.Est
  ) %>%
  dplyr::mutate(date = as.Date(date)) %>%
  dplyr::select(-report) %>%
  dplyr::mutate(time_diff = date - lag(date))

jumps <- which(MODEL3_CASES$time_diff > lubridate::make_difftime(day = 7))

temp <- MODEL3_CASES
for (j in 1:length(jumps)) {
  temp <- rbind(
    temp[1:(jumps[j] + j - 2), ],
    rep(NA, ncol(temp)),
    temp[(jumps[j] + j - 1):nrow(temp), ]
  )
}

@@

<<combineARMA_M3, echo=FALSE>>=
h3 <- haiti3_spatPomp()
deps <- h3@unit_names
ARMA_M3_joined <- dplyr::left_join(
  x = unitLikesM3,
  y = unitLikesARMA,
  by = "unit"
) %>%
  select(-ll_se) %>%
  mutate(unit = as.character(unit)) %>%
  pivot_longer(
    cols = -unit,
    names_to = "type",
    values_to = "ll"
  ) %>%
  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
@

@@

coef(h3)[names(best_params)] <- best_params

registerDoRNG(40765101)

bpf_out <- bake(
  file = "../model3/run_level_3/si_bpf.rds", {
    h3 %>% bpfilter(
      Np = 2000,
      block_size = 1,
      save_states = TRUE
    )
  }
)

@@

In an attempt to determine why the model is underperforming in some departments, we plot the conditional log likelihoods of each observation given the previous observations for each department in Fig.~\ref{fig:condLL}.
This figure reveals that the fitted model has difficult time explaining certain features of the data.
For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 that would increase the transmission between environmental cholera and humans for Sud and Grand'Anse.
The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.

\begin{figure}[ht]
<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
deps[deps == "Grande-Anse"] <- "Grand\'Anse"
names(deps) <- as.character(1:10)

myLabeller <- as_labeller(deps)

>>>>

<<packages, include=FALSE>>=
library(tidyverse)
library(pomp)
library(haitipkg)
library(spatPomp)
library(doParallel)
library(doRNG)

cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
if(is.na(cores)) cores <- detectCores()
# cores <- 20
registerDoParallel(cores)
theme_set(theme_bw())

if (!dir.exists("../model3/si")) {
  dir.create("../model3/si", recursive = TRUE)
}
@

<<set-opts, include=FALSE>>=
options(
  scipen = 2,
  help_type = "html",
  stringsAsFactors = FALSE,
  continue = "+  ",
  width = 70,
  useFancyQuotes = FALSE,
  reindent.spaces = 2,
  xtable.comment = FALSE
)
@

<<knitr-opts, include=FALSE, purl=FALSE>>=
library(knitr)
opts_knit$set(concordance=TRUE)
opts_chunk$set(
  progress=TRUE,prompt=TRUE,highlight=FALSE,
  tidy=TRUE,
  tidy.opts=list(
    keep.blank.line=FALSE
  ),
  comment="",
  warning=FALSE,
  message=FALSE,
  error=TRUE,
  echo=FALSE,
  cache=FALSE,
  strip.white=TRUE,
  results="markup",
  fig.path="figure/",
  fig.lp="fig:",
  fig.align="center",
  fig.show="asis",
  dev="pdf",
  dev.args=list(
    bg="transparent",
    pointsize=9
  )
)

@@

\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}

@@

In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
For a discrete population model, these rates define a Markov chain.
For a continuous and deterministic model, the rates define a system of ordinary differential equations.
Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
Our treatment follows \citet{breto09} \jwc{TODO: Cite the short course}.

A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
A flow into compartment $i$ from outside the system is denoted by $N_{\circ i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\circ}$.
We call $\circ$ a source/sink compartment, though it is an irregular compartment since $X_{circ}(t)$ is not defined.
These flows are required to satisfy a ``conservation of mass'' identity:
\begin{equation}
X_i(t)=X_i(t_0)+N_{\bullet i}(t) - N_{i\bullet}(t) + \sum_{j\neq
i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
\end{equation}
Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\bullet$.

@@

\subsection{Simple Markov counting system interpretation}
\label{subsec:smcs}
A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
A basic approach to this is to define
\begin{eqnarray}
\label{eq:smcs1}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
\\
\label{eq:smcs2}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
 &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
\end{eqnarray}
for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\bullet\}$ with $i\neq j$.
As with the ODE case, we need special attention for flows into the system, and we define
\begin{eqnarray}
\label{eq:smcs3}
\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta),
\\
\label{eq:smcs4}
\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=1 \given X(t)\big]
 &=& \delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta).
\end{eqnarray}
Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
We therefore call the Markov chain a simple Markov counting system (SMCS).
The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.

@@

Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
Specifically, we consider a model given by
\begin{equation}
\label{eq:odmcs1}
\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
\end{equation}
where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
\begin{equation}
\label{eq:odmcs2}
\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
\end{equation}
Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.

@@

Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
For the case $i=\bullet$, the treatment used by Model~1 is to set
\begin{equation}
\label{eq:numerics6}
\Delta N_{\bullet j} \sim \mathrm{poisson}( \mu_{\bullet j} \delta),
\end{equation}
an independent Poisson random variable with mean $\mu_{\bullet j} \delta$.

Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\bullet i}(t) = \sum_{i}N_{i \bullet}(t)$, in order to make the model consistent with the known total population.
In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\bullet$ into $S$ to describe only net population increase.

@@

This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
We therefore do not allow a change in reporting rate when fitting Model~3.

@@

To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values could be known, or could be a non-random function of the unknown parameter vector, $\theta$.
This density is used to obtain initial values of the latent state when fitting and evaluating the model.
For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
In the following subsections, we mention initial value parameters that were fit for each model.

@@

For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I(0)$ and $E(0)$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S(0) = 1 - I(0) - E(0)$.

@@

Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.

@@

As described in the main test, the latent states are initialized by estimating of the number of infected, asymptomatic, and recovered individuals at time $t_0$ by using observed case counts at time $t_1$.
The initial value for the bacteria compartment is then determined using the rainfall dynamics of Eq.~23.
For the departments of Grande'Anse, Nord-Ouest, and Sud-Est, estimating the latent states at time $t_0$ using the observed data at time $t_1$ is problematic (see Sec.~\ref{sec:mod3Cal}), so we instead estimate initial value of the latent state $I_{u0}(t_0)$ for the integers $u \in \{3, 7, 9\}$, which correspond to the aforementioned departments.
The remaining hidden states are estimated by enforcing model dynamics using this result.

@@

<<CalibrateOrig3>>=
set.seed(8636712)

# Create vectors for the unit and shared parameters
unit_specific_names <- c("betaB", "foi_add")
shared_param_names <- c(
  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
  "epsilon", "k"
)

# Vector containing all of the estimate parameter names.
est_param_names <- c(
  unit_specific_names, shared_param_names
)

@@

# This function sets the rw.sd for named parameters, first search.
set_rw_1 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.01)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.01)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.01)
  } else {
    return(0)
  }
}

# This function sets the rw.sd for named parameters, second search.
set_rw_2 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.003)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.003)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.003)
  } else {
    return(0)
  }
}

# This function sets the rw.sd for named parameters, final search.
set_rw_3 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.0025)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.0025)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.0025)
  } else {
    return(0)
  }
}

@@

# Search1 hyper-parameters
SEARCH1 = list(
  NBPF = 100,
  NP = 1000,
  SPAT_REGRESSION = 0.05,
  NREPS = 72,
  NP_EVAL = 1000,
  NREPS_EVAL = 8,
  RW_SD = chol_rw_1.sd,
  COOLING = 0.5,
  KEEP_TRACES = FALSE,
  KEEP_LIK_MAT = FALSE
)

# Search2 hyper-parameters
SEARCH2 = list(
  TOP_N = 9,
  NBPF = 100,
  NP = 1000,
  SPAT_REGRESSION = 0.05,
  NREPS = 4,
  NP_EVAL = 1000,
  NREPS_EVAL = 9,
  RW_SD = chol_rw_2.sd,
  COOLING = 0.5,
  KEEP_TRACES = FALSE,
  KEEP_LIK_MAT = FALSE
)

# Search3 hyper-parameters
SEARCH3 = list(
  TOP_N = 6,
  NBPF = 100,
  NP = 1000,
  SPAT_REGRESSION = 0.05,
  NREPS = 6,
  NP_EVAL = 1000,
  NREPS_EVAL = 9,
  RW_SD = chol_rw_3.sd,
  COOLING = 0.5,
  KEEP_TRACES = FALSE,
  KEEP_LIK_MAT = TRUE
)

@@

save_file <- "../model3/si/fit_orig_model.rds"

# Fit the model and save the results
haiti3_orig_fit <- bake(
  file = save_file, {
    fit_coupled_haiti3(
      nsearches = N_SEARCHES,
      search1 = SEARCH1,
      search2 = SEARCH2,
      search3 = SEARCH3,
      search_rho = FALSE,
      search_gamma = FALSE,
      ncores = cores
    )
  }
)

# Find model index with best log-likelihood
best_m <- which.max(haiti3_orig_fit$search3$logLiks$logLik)

# Find parameters using best_m index
best_params <- haiti3_orig_fit$search3$params[best_m, ]

# Get unit log-likelihoods for the best model.
all_liks <- as.data.frame(haiti3_orig_fit$search3$likMat)
all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
best_mod <- filter(all_liks, which == best_m)

# Pivot longer to get estimate of log-likelihood for each unit.
unitLikesM3 <- best_mod %>%
  pivot_longer(
    cols = -which,
    names_to = "unit",
    names_prefix = "V",
    values_to = "logLik"
  ) %>%
  select(-which) %>%
  group_by(unit) %>%
  summarize(
    ll_est = logmeanexp(logLik),
    ll_se = logmeanexp(logLik, se = TRUE)[2]
  ) %>%
  ungroup() %>%
  mutate(unit = as.numeric(unit)) %>%
  arrange(unit)
@

<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=

# Create dataset that will be used to fit ARMA models. To do this, we
# need to make sure that missing values are marked as NA, so we check
# for any jumps in time.
MODEL3_CASES <- haitiCholera %>%
  dplyr::rename(
    date = date_saturday, Grande_Anse = Grand.Anse,
    Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
    Sud_Est = Sud.Est
  ) %>%
  dplyr::mutate(date = as.Date(date)) %>%
  dplyr::select(-report) %>%
  dplyr::mutate(time_diff = date - lag(date))

# Find jumps
jumps <- which(MODEL3_CASES$time_diff > lubridate::make_difftime(day = 7))

# If there is a jump, insert NA between observations
temp <- MODEL3_CASES
for (j in 1:length(jumps)) {
  temp <- rbind(
    temp[1:(jumps[j] + j - 2), ],
    rep(NA, ncol(temp)),
    temp[(jumps[j] + j - 1):nrow(temp), ]
  )
}

@@

<<combineARMA_M3, echo=FALSE>>=
h3 <- haiti3_spatPomp()
deps <- h3@unit_names

# Combine tables with model 3 and benchmark likelihoods.
ARMA_M3_joined <- dplyr::left_join(
  x = unitLikesM3,
  y = unitLikesARMA,
  by = "unit"
) %>%
  select(-ll_se) %>%
  mutate(unit = as.character(unit)) %>%
  pivot_longer(
    cols = -unit,
    names_to = "type",
    values_to = "ll"
  ) %>%
  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
@

@@

# Change model parameters to the best set of parameters from previous
# search
coef(h3)[names(best_params)] <- best_params

registerDoRNG(40765101)  # for modular, reproducible results

# Perform block partical filter with fitted model, to get conditional
# log-likelihoods.
bpf_orig3_out <- bake(
  file = "../model3/si/bpf_orig3_cond_LL.rds", {

    bpf_out <- h3 %>% bpfilter(
      Np = 2000,
      block_size = 1,
      save_states = TRUE
    )

    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
      get_h3_S_quants(bpf_out, i)
    }

    results$block.cond.loglik <- bpf_out@block.cond.loglik
    results$times <- bpf_out@times  # Saved for convenience
    results
  },
  seed = 432235
)
@

@@

In an attempt to determine why the model is underperforming in some departments, we plot the conditional log likelihoods of each observation given the previous observations for each department in Fig.~\ref{fig:condLL}.
This figure reveals that the fitted model has difficult time explaining certain features of the data.
For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.

\begin{figure}[ht]
<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=

# Changing labels to human-readable labels.
deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
deps[deps == "Grande-Anse"] <- "Grand\'Anse"
names(deps) <- as.character(1:10)
myLabeller <- as_labeller(deps)
